---
title: "Houshold shortage"
author: "Joris heeremans(2787294), Tom Nethe(2784796), Mohamed Hannun(2824390), Jelle Menninga(2864828), Daan Notenboom (2853496), Suleyman Yigitsoy(2861891), Soufyan Attokoki() "
date: "`r Sys.Date()`"
output: pdf_document
---

# Set-up your environment

```{r package_install, eval=FALSE, include=FALSE}
install.packages("tidyverse")
devtools::install_github("r-lib/conflicted")
install.packages("cbsodataR")
install.packages("dplyr")
``

```{r packages}
require(tidyverse)
require(cbsodataR)
library(dplyr)
```

# Houshold Shortage

Joris Heeremans, Tom Nethe, Mohamed Hannun, Jelle Menninga, Daan Notenboom, Suleyman Yigitsoy, Soufyan Attokoki

Tutorial group number: 5

Tutorial lecturer's name: Jack Fitzgerald

# Part 1 - Identify a Social Problem

## 1.1 Describe the Social Problem

The Netherlands is dealing with a severe housing shortage, estimated at around 400,000 homes. (CBS, 2025) 
This imbalance between supply and demand, was created by population growth (including immigration), an increase in single-person households, 
and a lack of new construction (Langen, January 2025). This imbalance has led to skyrocketing prices and fierce competition for available houses. 
The house prices have risen significantly, making it difficult for many, especially solo buyers and those with low to middle incomes, 
to find suitable and affordable housing. The government needs to come up with new ideas, otherwise this problem will only grow bigger.



# Part 2 - Data Sourcing

## 2.1 Load in the data


```{r loading_data}
Voorraad_woningen <- read.csv("Vooraad_woningen_Google - Voorraad_woningen.csv")
KW1_voorraad <- Voorraad_woningen[grepl("1e kwartaal", Voorraad_woningen$Perioden), ]

Huishoudens <- read.csv("Aantal_huishoudens - Blad1 (2).csv")

```


## 2.2 Provide a short summary of the dataset(s)

```{r}
head(KW1_voorraad)
head(Huishoudens)
```

Both data sets include information about the stock houses and the number of households in the Netherlands per province and period. 


# Part 3 - Quantifying

## 3.1 Data cleaning

We already filtered the data through the CBS filter.  
```{r data_cleaning}


# Add a year column by extracting the first 4 characters of the period column
Voorraad_woningen$Year <- substr(Voorraad_woningen$Perioden, 1, 4)
Huishoudens$Year <- substr(Huishoudens$Periode, 1, 4)

# Set which years we want to use
years_to_keep <- c("2019", "2020", "2021")

#Set which provinces 
provincies <- c("Groningen (PV)", "Friesland (PV)", "Drenthe (PV)", "Overijssel (PV)", "Flevoland (PV)",
                "Gelderland (PV)", "Utrecht (PV)", "Noord-Holland (PV)", "Zuid-Holland (PV)",
                "Zeeland (PV)", "Noord-Brabant (PV)", "Limburg (PV)")

# Select only data for Netherlands or provinces, and only the needed years, only first quarter
housing_stock_selected <- subset(Voorraad_woningen,
                                 Regio.s %in% provincies &
                                 Year %in% years_to_keep &
                                 grepl("1e kwartaal", Perioden))


households_selected <- subset(Huishoudens,
                               Regio.s %in% provincies &
                               Year %in% years_to_keep)



```

## 3.2 Generate necessary variables

# Merge datasets
data_combined <- housing_stock_selected %>%
  select(Regio.s, Year, Voorraad = aantal) %>%  # Replace value_column if needed
  left_join(
    households_selected %>%
      select(Regio.s, Year, Huishoudens = aantal), # Replace value_column if needed
    by = c("Regio.s", "Year")
  )

# Create shortage variable
data_combined <- data_combined %>%
  mutate(Tekort = Huishoudens - Voorraad)
```

Variable 2

```{r gen_var2}

```

## 3.3 Visualize temporal variation

```{r}

```

## 3.4 Visualize spatial variation

```{r visualise_map}

```

Here you provide a description of why the plot above is relevant to your specific social problem.

## 3.5 Visualize sub-population variation

What is the poverty rate by state?

```{r visualise_subpopulations}
dataset$inmetro <- dataset$inmetro %>% as.factor()
# Boxplot of poverty rate by state using the 'midwest' dataset
ggplot(dataset, aes(x = inmetro, y = percadultpoverty)) +
  geom_boxplot() +
  labs(
    title = "Distribution of Poverty Rates by Metropolitan status (Midwest counties)",
    x = "Metropolitan Area",
    y = "Poverty Rate of Adults (%)"
  ) +
  theme_minimal() +
  theme(
    legend.position = "right"
  )
```
library(dplyr)
library(ggplot2)

voorraad_nl <- KW1_voorraad %>% filter(Regio.s == "Nederland")
huishoudens_nl <- Huishoudens %>% filter(Regio.s == "Nederland")

colnames(voorraad_nl) <- c("Regio", "Periode", "Voorraad")
colnames(huishoudens_nl) <- c("Regio", "Periode", "Huishoudens")


combined <- merge(voorraad_nl, huishoudens_nl, by = c("Regio", "Periode"))

# Maak de data 'long' voor ggplot
combined_long <- combined %>%
  select(Periode, Voorraad, Huishoudens) %>%
  tidyr::pivot_longer(cols = c("Voorraad", "Huishoudens"),
                      names_to = "Type",
                      values_to = "Aantal")

ggplot(combined_long, aes(x = Periode, y = Aantal, color = Type, group = Type)) +
  geom_line(size = 1.2) +
  labs(title = "Vergelijking woningvoorraad en huishoudens (Nederland)",
       x = "Periode", y = "Aantal", color = "black") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

Here you provide a description of why the plot above is relevant to your specific social problem.

## 3.6 Event analysis

Analyze the relationship between two variables.

```{r analysis}

```

Here you provide a description of why the plot above is relevant to your specific social problem.

# Part 4 - Discussion

## 4.1 Discuss your findings

# Part 5 - Reproducibility

## 5.1 Github repository link

https://github.com/ProgrammingforEcon-Team-5-Groep-1/Programming-Group5-Team1

## 5.2 Reference list
CBS Statline. (May 23, 2025). https://opendata.cbs.nl/statline/#/CBS/nl/dataset/81955NED/table?fromstatweb (Last used on June 2, 2025)
CBS Statline. (June 3, 2025). https://opendata.cbs.nl/statline/#/CBS/nl/dataset/71486ned/table?fromstatweb (Last used on June 2, 2025)
Centraal Bureau voor de Statistiek. (2025). Wonen. Centraal Bureau Voor de Statistiek. https://www.cbs.nl/nl-nl/visualisaties/monitor-brede-welvaart-en-de-sustainable-development-goals/hier-en-nu/wonen#:~:text=Door%20ABF%20Research%20is%20voor,een%20kleine%20aanvullende%20aardgaslevering%20plaatsvindt.
Mike Langen, ABN AMRO Bank. (16 January, 2025). Housing market - building according to need. https://www.abnamro.com/research/en/our-research/housing-market-building-according-to-need

Use APA referencing throughout your document.
