---
title: "Houshold shortage"
author: "Joris heeremans(2787294), Tom Nethe(2784796), Mohamed Hannun(2824390), Jelle Menninga(2864828), Daan Notenboom (2853496), Suleyman Yigitsoy(2861891), Soufyan Attokoki() "
date: "`r Sys.Date()`"
output: pdf_document
---

# Set-up your environment

```{r package_install, eval=FALSE, include=FALSE}
install.packages("tidyverse")
devtools::install_github("r-lib/conflicted")
install.packages("cbsodataR")
install.packages("dplyr")
install.packages("sf")
install.packages("rnaturalearth")
install.packages("rnaturalearthdata")
```
```{r packages, include=FALSE}
require(tidyverse)
require(cbsodataR)
library(dplyr)
library(ggplot2)
library(sf)
library(rnaturalearth)
library(rnaturalearthdata)
library(dplyr)
library(ggplot2)

```

# Houshold Shortage

Joris Heeremans, Tom Nethe, Mohamed Hannun, Jelle Menninga, Daan Notenboom, Suleyman Yigitsoy, Soufyan Attokoki

Tutorial group number: 5

Tutorial lecturer's name: Jack Fitzgerald

# Part 1 - Identify a Social Problem

## 1.1 Describe the Social Problem

The Netherlands is dealing with a severe housing shortage, estimated at around 400,000 homes. (CBS, 2025) 
This imbalance between supply and demand, was created by population growth (including immigration), an increase in single-person households, 
and a lack of new construction (Langen, January 2025). This imbalance has led to skyrocketing prices and fierce competition for available houses. 
The house prices have risen significantly, making it difficult for many, especially solo buyers and those with low to middle incomes, 
to find suitable and affordable housing. The government needs to come up with new ideas, otherwise this problem will only grow bigger.



# Part 2 - Data Sourcing

## 2.1 Load in the data


```{r loading_data}
Voorraad_woningen <- read.csv("Vooraad_woningen_Google - Voorraad_woningen.csv")
KW1_voorraad <- Voorraad_woningen[grepl("1e kwartaal", Voorraad_woningen$Perioden), ]

Huishoudens <- read.csv("Aantal_huishoudens - Blad1 (2).csv")

Migratie <- read.csv("Migratie - Blad1 (2).csv")

Bevolking <- read.csv("Bevolking - Blad1.csv")

```


## 2.2 Provide a short summary of the dataset(s)

```{r}
head(KW1_voorraad)
head(Huishoudens)
```

Both data sets include information about the stock houses and the number of households in the Netherlands per province and period. 


# Part 3 - Quantifying

## 3.1 Data cleaning

We already filtered the data through the CBS filter.  
```{r data_cleaning}


# Add a year column by extracting the first 4 characters of the period column
Voorraad_woningen$Year <- substr(Voorraad_woningen$Perioden, 1, 4)
Huishoudens$Year <- substr(Huishoudens$Periode, 1, 4)


# Set which years we want to use
years_to_keep <- c("2019", "2020", "2021")

#Set which provinces 
provincies <- c("Groningen (PV)", "Fryslân (PV)", "Drenthe (PV)", "Overijssel (PV)", "Flevoland (PV)",
                "Gelderland (PV)", "Utrecht (PV)", "Noord-Holland (PV)", "Zuid-Holland (PV)",
                "Zeeland (PV)", "Noord-Brabant (PV)", "Limburg (PV)")

# Select only data for Netherlands or provinces, and only the needed years, only first quarter
housing_stock_selected <- subset(Voorraad_woningen,
                                 Regio.s %in% provincies &
                                 Year %in% years_to_keep &
                                 grepl("1e kwartaal", Perioden))


households_selected <- subset(Huishoudens,
                               Regio.s %in% provincies &
                               Year %in% years_to_keep)

# --- 1. Reshape migration data from wide to long ---
migratie_long <- Migratie %>%
  filter(Onderwerp %in% c("Immigratie", "Emigratie")) %>%
  select(-X) %>%  # Verwijder de kolom die problemen veroorzaakt
  pivot_longer(
    cols = starts_with("X"),
    names_to = "Year",
    names_prefix = "X",
    values_to = "Aantal"
  ) %>%
  select(Regio.s, Onderwerp, Year, Aantal) %>%
  pivot_wider(
    names_from = Onderwerp,
    values_from = Aantal
  ) %>%
  mutate(
    Immigratie = as.numeric(Immigratie),
    Emigratie = as.numeric(Emigratie)
  )
# --- 2. Reshape population data from wide to long ---
bevolking_long <- Bevolking %>%
  filter(Onderwerp == "Bevolking") %>%
  select(-X) %>%  # Verwijder de kolom die de fout veroorzaakt
  pivot_longer(
    cols = starts_with("X"),
    names_to = "Year",
    names_prefix = "X",
    values_to = "Bevolking"
  ) %>%
  select(Regio.s, Year, Bevolking) %>%
  mutate(Bevolking = as.numeric(Bevolking))

# --- 3. Merge and calculate net migration as % of population ---
migratie_bevolking <- migratie_long %>%
  left_join(bevolking_long, by = c("Regio.s", "Year")) %>%
  mutate(
    Netto_Migratie = Immigratie - Emigratie,
    Netto_Migratie_Perc = 100 * Netto_Migratie / Bevolking
  )

head(migratie_bevolking)


```

```{r gen_Var}

## 3.2 Generate necessary variables

# Merge datasets
data_combined <- housing_stock_selected %>%
  select(Regio.s, Year, Voorraad = aantal) %>%  # Replace value_column if needed
  left_join(
    households_selected %>%
      select(Regio.s, Year, Huishoudens = aantal), # Replace value_column if needed
    by = c("Regio.s", "Year")
  )

# Create shortage variable
data_combined <- data_combined %>%
  mutate(Tekort = Huishoudens - Voorraad)


```

```{r gen_ar}
# SUM the shortage across all provinces for each year
nederland_shortage <- data_combined %>%
  group_by(Year) %>%
  summarize(Total_Tekort = sum(Tekort, na.rm = TRUE))

# Zorg dat Year een factor is
nederland_shortage$Year <- as.factor(nederland_shortage$Year)

# Maak de lijnplot met vaste y-as (0 tot 100.000)
ggplot(nederland_shortage, aes(x = Year, y = Total_Tekort, group = 1)) +
  geom_line(color = "black", size = 1.2) +
  geom_point(size = 3, color = "red") +
  scale_y_continuous(
    limits = c(0, 100000),
    breaks = seq(0, 100000, by = 20000),
    labels = scales::comma
  ) +
  labs(
    title = "Housing Shortage in the Netherlands (2019–2021)",
    x = "Year",
    y = "Housing Shortage"
  ) +
  theme_minimal(base_size = 13)

``` 
``` {r gen_var2}
## 3.3 Visualize temporal variation


# Zorg dat je factor-niveau voor jaar logisch wordt geordend (voor de x-as)
data_combined$Year <- as.factor(data_combined$Year)

# zwarte rand, smallere staven optioneel
ggplot(data_combined, aes(x = factor(Year), y = Tekort, fill = Regio.s)) +
  geom_col(position = "dodge", color = "black", width = 0.8) +
  scale_y_continuous(
    breaks = seq(
      floor(min(data_combined$Tekort, na.rm = TRUE) / 2000) * 2000, 
      ceiling(max(data_combined$Tekort, na.rm = TRUE) / 2000) * 2000, 
      by = 2000)
  ) +
  labs(
    title = "Housing Shortage per year per province",
    x = "Year",
    y = "Housing Shortage",
    fill = "Province"
  ) +
  theme_minimal()


```

## 3.4 Visualize spatial variation

```{r visualise_map}
# Haal shapefile met Nederlandse provincies binnen
nl_prov <- ne_states(country = "Netherlands", returnclass = "sf")

# Selecteer alleen de 12 echte provincies van Nederland
provincies_nederland <- c(
  "Drenthe", "Flevoland", "Friesland", "Gelderland", "Groningen",
  "Limburg", "Noord-Brabant", "Noord-Holland", "Overijssel",
  "Utrecht", "Zeeland", "Zuid-Holland"
)
nl_prov_alleen_prov <- nl_prov %>% filter(name %in% provincies_nederland)

# Maak een data-frame voor 2021 met provincie-kolom zonder (PV)
data_2021 <- data_combined %>%
  filter(Year == "2021") %>%
  mutate(provincie_kort = gsub(" \\(PV\\)", "", Regio.s),
         provincie_kort = ifelse(provincie_kort == "Fryslân", "Friesland", provincie_kort))

# Join: voeg de tekorten toe aan de kaart
kaart_met_tekort <- nl_prov_alleen_prov %>%
  left_join(data_2021, by = c("name" = "provincie_kort"))

# Plot de kaart met kleurverloop blauw (laag) tot rood (hoog)
ggplot(kaart_met_tekort) +
  geom_sf(aes(fill = Tekort), color = "black", size = 0.5) +
  scale_fill_gradient(low = "white", high = "red", name = "Woningtekort") +
  labs(
    title = "Housing shortage per province (2021)",
    caption = "Bron: CBS & eigen analyse"
  ) +
  theme_minimal(base_size = 10) +
  theme(
    legend.key.height = unit(1.5, "cm"),
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    panel.grid = element_blank())

```

Here you provide a description of why the plot above is relevant to your specific social problem.

## 3.5 Visualize sub-population variation

What is the poverty rate by state?

```{r visualise_subpopulations}
dataset$inmetro <- dataset$inmetro %>% as.factor()
# Boxplot of poverty rate by state using the 'midwest' dataset
ggplot(dataset, aes(x = inmetro, y = percadultpoverty)) +
  geom_boxplot() +
  labs(
    title = "Distribution of Poverty Rates by Metropolitan status (Midwest counties)",
    x = "Metropolitan Area",
    y = "Poverty Rate of Adults (%)"
  ) +
  theme_minimal() +
  theme(
    legend.position = "right"
  )
```

  theme(axis.text.x = element_text(angle = 45, hjust = 1))

Here you provide a description of why the plot above is relevant to your specific social problem.

## 3.6 Event analysis

Analyze the relationship between two variables.

```{r analysis}

```

Here you provide a description of why the plot above is relevant to your specific social problem.

# Part 4 - Discussion

## 4.1 Discuss your findings

# Part 5 - Reproducibility

## 5.1 Github repository link

https://github.com/ProgrammingforEcon-Team-5-Groep-1/Programming-Group5-Team1

## 5.2 Reference list

CBS Statline. (May 23, 2025). https://opendata.cbs.nl/statline/#/CBS/nl/dataset/81955NED/table?fromstatweb (Last used on June 2, 2025)
CBS Statline. (June 3, 2025). https://opendata.cbs.nl/statline/#/CBS/nl/dataset/71486ned/table?fromstatweb (Last used on June 2, 2025)
CBS Statline. (May 28, 2025). https://opendata.cbs.nl/statline/#/CBS/nl/dataset/37230ned/table?ts=1750160957830 (Last used on June 17,2025)
CBS Statline. (May 28, 2025). https://opendata.cbs.nl/statline/#/CBS/nl/dataset/37230ned/table?ts=1750159602778 (Last used on June 17,2025)
Centraal Bureau voor de Statistiek. (2025). Wonen. Centraal Bureau Voor de Statistiek. https://www.cbs.nl/nl-nl/visualisaties/monitor-brede-welvaart-en-de-sustainable-development-goals/hier-en-nu/wonen#:~:text=Door%20ABF%20Research%20is%20voor,een%20kleine%20aanvullende%20aardgaslevering%20plaatsvindt.
Mike Langen, ABN AMRO Bank. (16 January, 2025). Housing market - building according to need. https://www.abnamro.com/research/en/our-research/housing-market-building-according-to-need

Use APA referencing throughout your document.
